---
title: Applied Machine Learning for Educational Data Science
subtitle: Introduction to Toy Datasets
author:
  name: Cengiz Zopluoglu
  affiliation: University of Oregon | EDLD 654
date: 06/28/2021 ## Or "Lecture no."
output: 
  html_document:
    keep_md: false
    theme: journal
    highlight: haddock
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    extra_dependencies: ["amssymb","animate"]
    keep_tex: false ## Change to true if want keep intermediate .tex file
    toc: true
    toc_depth: 3
    dev: cairo_pdf
always_allow_html: true
urlcolor: blue
mainfont: cochineal
sansfont: Fira Sans
monofont: Fira Code ## Although, see: https://tex.stackexchange.com/q/294362

## Automatically knit to both formats:
knit: (function(inputFile, encoding) {
 rmarkdown::render(inputFile, encoding = encoding, 
 output_format = 'all') 
 })
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position=c('top','right'))
```

<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    z-index: 2;
    color: #fff;
    background-color: #FC4445;
    border-color: #97CAEF;
}

</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "",fig.align='center')
require(here)
require(ggplot2)
require(plot3D)
require(kableExtra)
require(knitr)
require(giski)
require(magick)
options(scipen=99)

```

`r paste('[Updated:',format(Sys.time(),'%a, %b %d, %Y - %H:%M:%S'),']')`

There are two datasets we will analyze throughout the whole course. The first dataset has a continuous outcome and the second dataset has a binary outcome. We will apply several methods and algorithms to these two datasets during the course. This will give us an opportunity to compare and contrast the prediction outcomes from several models and methods on the same datasets. This section provides some background information and context for these two datasets.

# Readability

The readability dataset comes from a recent [Kaggle Challange (CommonLit Readability Prize)](https://www.kaggle.com/c/commonlitreadabilityprize/). You can directly download the training dataset from the competition website, or download it from [this link](https://github.com/uo-datasci-specialization/c4-ml-fall-2021/tree/main/data)


[Download the item parameter file](https://github.com/czopluoglu/website/blob/master/docs/posts/2020-12-04-detect/data/ipar.csv)



# Recidivism













