---
title: K-Nearest Neighbors
subtitle: Applied Machine Learning for Educational Data Science
author:
  name: Cengiz Zopluoglu
  affiliation: University of Oregon | EDLD 654
date: 11/01/2021 ## Or "Lecture no."
output: 
  html_document:
    keep_md: false
    theme: journal
    highlight: haddock
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    extra_dependencies: ["amssymb","animate","upgreek","amsmath"]
    keep_tex: false ## Change to true if want keep intermediate .tex file
    toc: true
    toc_depth: 3
    dev: cairo_pdf
always_allow_html: true
urlcolor: blue
mainfont: cochineal
sansfont: Fira Sans
monofont: Fira Code ## Although, see: https://tex.stackexchange.com/q/294362

## Automatically knit to both formats:
knit: (function(inputFile, encoding) {
 rmarkdown::render(inputFile, encoding = encoding, 
 output_format = 'all') 
 })
---

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position=c('top','right'))
```

`r paste('[Updated:',format(Sys.time(),'%a, %b %d, %Y - %H:%M:%S'),']')`

<style>
  
  .list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    z-index: 2;
    color: #fff;
      background-color: #FC4445;
      border-color: #97CAEF;
  }

#infobox {
padding: 1em 1em 1em 4em;
margin-bottom: 10px;
border: 2px solid black;
border-radius: 10px;
background: #E6F6DC 5px center/3em no-repeat;
  }

</style>
  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "",fig.align='center')
require(here)
require(ggplot2)
require(plot3D)
require(kableExtra)
require(knitr)
require(giski)
require(magick)
require(gridExtra)
require(plotly)
options(scipen=99)


# Resources:
# https://bradleyboehmke.github.io/HOML/knn.html
# Applied Predictive Modeling, Chapter 13.5
# https://towardsdatascience.com/machine-learning-basics-with-the-k-nearest-neighbors-algorithm-6a6e71d01761
#https://towardsdatascience.com/a-simple-introduction-to-k-nearest-neighbors-algorithm-b3519ed98e
```

# 1. K-Nearest Neighbors for Continuous Outcomes


# 2. K-Nearest Neighbors for Binary Outcomes


